Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-02-07 parallel-tractability-J-v9/parallelTractabilityOntMat.tex
overall revision to make some presentations consistent

In this section, we study the issue of parallel tractability
for materialization of DL-Lite and DHL (DHL($\circ$)) ontologies
based on Algorithm~$\mathsf{A}_{\text{opt}}$. We show that, for any DL-Lite$_{\text{core}}$ or DL-Lite$_\mathcal{R}$ ontology $\mathcal{O}$,
there exists a poly-logarithmically bounded function $\psi$
such that Algorithm~$\mathsf{A}_{\text{opt}}^{\psi}$ can handle materialization of $\mathcal{O}$.
However, for DHL and DHL($\circ$), there exist ontology classes such that Algorithm~$\mathsf{A}_{\text{opt}}^\psi$
does not work. We illustrate the reason why Algorithm~$\mathsf{A}_{\text{opt}}^\psi$ cannot always
work by studying specific cases.
Further, we propose to restrict the usage of DHL and DHL($\circ$) in order to achieve parallel tractability
of materialization.

@@

In this part, we study materialization tractable in parallel for DHL($\circ$) ontologies.
In addition to the rules in DHL, we also have to consider complex RIAs (R4).
We next show that complex RIAs may also cause the situation of `path twisting'.
Consider the following example:
%
\begin{example}\label{exp:complexRIA}
Given a DHL$(\circ)$ ontology $\mathcal{O}_{\text{ex}_9}$ where the TBox is empty;
the RBox $\mathcal{R}$ contains three axioms:
$R_1\circ R_2\sqsubseteq R$, $R_3\circ R\sqsubseteq R_1$ and $R\circ R_4\sqsubseteq R_2$;
the ABox $\mathcal{A}$ is $\{R(a_1,a_1), R_3(a_i,a_{i-1}), R_4(a_{i-1},a_i)\}$
for $2\leq i\leq k$ and $k$ an integer greater than $2$.
The corresponding datalog program $P_{\text{ex}_5}$
contains three rules: `$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$',
`$R_3(x,y),R(y,z)\rightarrow R_1(x,z)$' and `$R(x,y),R_4(y,z)\rightarrow R_2(x,z)$'.
The materialization graph of $P_{\text{ex}_5}$ constructed by Algorithm~$\mathsf{A}_{\text{opt}}$ is denoted by $\mathcal{G}_{\text{ex}_9}$.
\end{example}

@@

\begin{example}
For the ontology $\mathcal{O}_{\text{ex}_7}$ in Example~\ref{exp:complexRIA}, all of the roles
$R_1$, $R_2$ and $R$ are non-simple roles. Thus, $\mathcal{O}_{\text{ex}_7}$ does not follow
the simple-role restriction because of $R_1\circ R_2\sqsubseteq R$.
Consider the ontology $\mathcal{O}_{\text{ex}_1}$
in Example~\ref{exp:mg} again. The role $R$ is a non-simple role, while $S$ is a simple role.
Thus, $\mathcal{O}_{\text{ex}_1}$ follows the simple-role restriction. All the implicit nodes
in $\mathcal{G}_{\text{ex}_1}$ have corresponding SWD paths in the first iteration.
Thus, `path twisting' cannot occur when materializing $\mathcal{O}_{\text{ex}_1}$ by Algorithm~$\mathsf{A}_{\text{opt}}$.
\end{example}

>>>>

In this section, we study the issue of parallel tractability
for materialization of DL-Lite and DHL (DHL($\circ$)) ontologies
based on Algorithm~$\mathsf{A}_{\text{opt}}$. We show that, for any class $\mathbb{O}$ of
DL-Lite$_{\text{core}}$ or DL-Lite$_\mathcal{R}$ ontologies,
there exists a poly-logarithmically bounded function $\psi$
such that Algorithm~$\mathsf{A}_{\text{opt}}^{\psi}$ can handle materialization of the ontologies in $\mathbb{O}$.
However, for DHL and DHL($\circ$), there exist ontology classes such that Algorithm~$\mathsf{A}_{\text{opt}}^\psi$
does not work. We illustrate the reason why Algorithm~$\mathsf{A}_{\text{opt}}^\psi$ cannot always
work by studying specific cases.
Further, we propose to restrict the usage of DHL and DHL($\circ$) in order to achieve parallel tractability
of materialization.

@@

In this part, we study materialization tractable in parallel for DHL($\circ$) ontologies.
In addition to the rules in DHL, we also have to consider complex RIAs (R4).
We next show that complex RIAs may also cause the situation of `path twisting'.
Consider the following example:
%
\begin{example}\label{exp:complexRIA}
Given a DHL$(\circ)$ ontology $\mathcal{O}_{\text{ex}_9}$ where the TBox is empty;
the RBox $\mathcal{R}$ contains three axioms:
$R_1\circ R_2\sqsubseteq R$, $R_3\circ R\sqsubseteq R_1$ and $R\circ R_4\sqsubseteq R_2$;
the ABox $\mathcal{A}$ is $\{R(a_1,a_1), R_3(a_i,a_{i-1}), R_4(a_{i-1},a_i)\}$
for $2\leq i\leq k$ and $k$ an integer greater than $2$.
The corresponding datalog program $P_{\text{ex}_9}$
contains three rules: `$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$',
`$R_3(x,y),R(y,z)\rightarrow R_1(x,z)$' and `$R(x,y),R_4(y,z)\rightarrow R_2(x,z)$'.
The materialization graph of $P_{\text{ex}_9}$ constructed by Algorithm~$\mathsf{A}_{\text{opt}}$ is denoted by $\mathcal{G}_{\text{ex}_9}$.
\end{example}

@@

\begin{example}
For the ontology $\mathcal{O}_{\text{ex}_9}$ in Example~\ref{exp:complexRIA}, all of the roles
$R_1$, $R_2$ and $R$ are non-simple roles. Thus, $\mathcal{O}_{\text{ex}_9}$ does not follow
the simple-role restriction because of $R_1\circ R_2\sqsubseteq R$.
Consider the ontology $\mathcal{O}_{\text{ex}_1}$
in Example~\ref{exp:mg} again. The role $R$ is a non-simple role, while $S$ is a simple role.
Thus, $\mathcal{O}_{\text{ex}_1}$ follows the simple-role restriction. All the implicit nodes
in $\mathcal{G}_{\text{ex}_1}$ have corresponding SWD paths in the first iteration.
Thus, `path twisting' cannot occur when materializing $\mathcal{O}_{\text{ex}_1}$ by Algorithm~$\mathsf{A}_{\text{opt}}$.
\end{example}

### 2018-02-07 parallel-tractability-J-v9/relatedWork.tex
overall revision to make some presentations consistent

The parallel reasoner RDFox \cite{MotikNPHO14} is used to evaluate our implementation. RDFox is a state-of-the-art system that handles reasoning on datalog rewritable ontology languages. Algorithm~$\mathsf{A}_{bsc}$ proposed in Section~\ref{sec:ptclass} is similar to the main algorithm of RDFox (see \cite{MotikNPHO14}, Sections~3 and 4). The difference lies in that, a group of rule instantiations are handled by one processor (namely a thread) in RDFox, while in Algorithm~$\mathsf{A}_{bsc}$, each rule instantiation is assigned to a unique processor. Thus, in theory, the materialization of the datalog program in Example~\ref{exp:mg} is serial on RDFox. It is also shown from the experiments that parallelism does not lead to a remarkable improvement when RDFox handles the ontology series in $\mathcal{D}^-_{\textit{\text{dhl}}(\circ)}$. We use Algorithm~$\mathsf{A}_{opt}$ to show that the ontology in Example~\ref{exp:mg} is also parallelly tractable. Our implementation ParallelDHL, which is based on Algorithm~$\mathsf{A}_{prc}$, has also been shown to have a better performance than RDFox when handling the test ontologies that belong to $\mathcal{D}_{\textit{\text{dhl}}(\circ)}$.

>>>>

The parallel reasoner RDFox \cite{MotikNPHO14} is used to evaluate our implementation. RDFox is a state-of-the-art system that handles reasoning on datalog rewritable ontology languages. Algorithm~$\mathsf{A}_{bsc}$ proposed in Section~\ref{sec:ptclass} is similar to the main algorithm of RDFox (see \cite{MotikNPHO14}, Sections~3 and 4). The difference lies in that, a group of rule instantiations are handled by one processor (namely a thread) in RDFox, while in Algorithm~$\mathsf{A}_{bsc}$, each rule instantiation is assigned to a unique processor. Analogous to the method used in RDFox, ParallelDHL is also implemented by assigning a group of 
rule instantiations to a processor.
The experimental results show that ParallelDHL is comparable to RDFox when handling
the LUBM and the YAGO ontologies.
